---
title: "Building standalone Rmarkdown documents using the data language engine"
author: "David M. Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
params:
  run_gpg: FALSE
vignette: >
  %\VignetteIndexEntry{Building standalone Rmarkdown documents using the data language engine}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(knitrdata) # load package

# Change document root to not write to user or package space
td = tempdir(check=TRUE)
knitr::opts_knit$set(root.dir = td)
```

This document provides a brief introduction to how to use the `knitrdata` package to create standalone Rmarkdown documents. For examples of Rmarkdown documents using the package, please consult the `examples` directory provided with the package. 

## Conceptual overview

Sometimes it would be useful to make completely standalone Rmarkdown documents that do not depend on data in external files. One important example of this is scientific publications written in Rmarkdown for which we often would like to supply the source document with the data to ensure results are reproducible. The `knitrdata` package addresses this need by creating a mechanism for incorporating arbitrary text and binary data in Rmarkdown documents. It works conceptually and technically in a manner that is very similar to how images and other binary data are incorporated into standalone HTML web pages: data are incorporated into specially delimited chunks that consist of the data themselves plus a small bit of header information explaining how the data are to be processed. Text data (e.g., CSV data tables, BibTeX references, LaTeX style files) is typically incorporated in chunks as is, whereas binary data (e.g., RDS files, images, NetCDF files) is encoded as text using one of two standard encoding schemes. During knitting of the Rmarkdown document, chunk data is decoded if necessary, after which it can either be loaded into the Rmarkdown R session or saved to an external file.

`knitrdata` achieves this by extending `knitr` to provide a new `data` language engine. Instead of putting code inside `data` chunks, one puts the contents of the data file that one wishes to use in your Rmarkdown document. For binary data, the package currently supports two standard encoding formats: `base64`, the standard binary encoding format used (behind the scenes) for things like email attachments and standalone HTML web pages; and `gpg`, a well-known encryption algorithm that prevents data from being accessed by users without the appropriate decryption key. The latter option requires that a GPG keyring for managing encryption keys be installed and properly configured.

`data` chunks do not produce output in the form of text or figures as most code chunks do. Instead, the decoded contents of the chunk are either returned as a variable in the R workspace or saved to an external file.

## Installation & getting started

`knitrdata` can be installed from its [github repository](https://github.com/dmkaplan2000/knitrdata) using the package [remotes](https://cran.r-project.org/package=remotes):

```{r eval=FALSE}
remotes::install_github("dmkaplan2000/knitrdata",build_vignettes=TRUE)
```

Once the package is installed, it needs to be loaded in the Rmarkdown script before the first `data` chunk, typically in the `setup` chunk at the start of the document:

```{r eval=FALSE}
library(knitrdata)
```

To ensure that your document is as standalone as possible, you can combine these two steps in the `setup` chunk (though this may install packages without asking the user):

```{r eval=FALSE}
# If package not installed, install it
if (!requireNamespace("knitrdata")) {
  if (!requireNamespace("remotes"))
    install.packages("remotes")
  remotes::install_github("dmkaplan2000/knitrdata",build_vignettes = TRUE)  
}

library(knitrdata) # load package
```

After the package is installed, `data` chunks can be incorporated in the document. The precise format for `data` chunks is described in the ["Data chunks"](#chunks) Section, but beforehand encoding of binary data is presented. If you are only interested in incorporating text data in your Rmarkdown document, then you can safely skip directly to the ["Data chunks"](#chunks) Section.

## Encoding data {#encode}

Both text and binary data files can be encoded, but encoding is only required or regularly used for binary data. Two encoding formats are currently implemented: `base64`, used for non-sensitive data; and `gpg`, allowing one to encrypt data so that only users with the decryption key have access. The latter option requires that a GPG keyring be installed and properly configured.

Two helper functions, `data_encode` and `data_decode`, are included in the package to facilitate encoding and decoding of data files. These are basically wrapper functions for functionality provided by the [base64enc](https://cran.r-project.org/package=base64enc) and [gpg](https://cran.r-project.org/package=gpg) packages. To demonstrate their use, we will use the following simple data frame that exists in both text (CSV) and binary (RDS) formats:

```{r}
D = data.frame(a=1:3,b=letters[1:3])
write.csv(D,"test.csv", row.names = FALSE, quote = FALSE)
saveRDS(D,"test.RDS")
D
```

The contents of the CSV file are as follows:

```{r comment="",echo=FALSE}
cat(readLines("test.csv"),sep="\n")
```

This CSV text will be used as is in `data` chunks (see ["Data chunks"](#chunks) Section for details).

### Base64

Base64 encoding is the standard encoding to be used for all non-sensitive binary data. It is based on translating 6-bits of information into one of 64 alphanumeric and symbolic characters.

Encoding in base64 using the `data_encode` function works as follows:

```{r}
b64 = knitrdata::data_encode("test.RDS","base64")
```

By default this function will silently return the encoded data as a character string. This character string can then be visualized using the `cat` function so that it can be copied and pasted directly into a `data` chunk:

```{r}
cat(b64)
```

This is only practical for relatively small data files, so for larger files, one can place the output in a file:

```{r}
data_encode("test.RDS","base64",output="test.RDS.base64")
cat(readLines("test.RDS.base64"),sep="\n")
```

Though it is rarely necessary to call the `data_decode` function directly when working with the `data` chunks in Rmarkdown documents, bas64 encoded data can be decoded as follows:

```{r}
rds = data_decode(b64,"base64",as_text=FALSE)
writeBin(rds,"test_output.RDS")
y = readRDS("test_output.RDS")
y
```

### GPG

Ecryption of data using GPG requires a properly configured GPG keyring. The functioning of GPG and GPG keyrings is beyond the scope of this document, but numerous websites explain how GPG works and how to install a GPG keyring, including the main [gpg](https://gnupg.org/) website.

For the purposes of this vignette, I will generate a test GPG private-public key pair using the [gpg](https://cran.r-project.org/package=gpg) package, however in real use scenarios proper keys would typically be generated using the `gpg` command line tool (or an equivalent alternative) with appropriate options.

```{r eval=requireNamespace("gpg") && params$run_gpg}
id = gpg::gpg_keygen("test","test@test.org")
```

Next one uses this key to encode a data file:

```{r eval=requireNamespace("gpg") && params$run_gpg}
enc = data_encode("test.RDS","gpg",options = list(receiver=id))
cat(enc)
```

Note that the ID of the desired encryption key must be supplied as the `receiver` in the `options` list input argument.

Decoding works as follows:

```{r eval=requireNamespace("gpg") && params$run_gpg}
rds = data_decode(enc,"gpg")
writeBin(rds,"test_output.RDS")
y = readRDS("test_output.RDS")
y
```

Note that there is no need to supply the receiver ID when decoding because the appopriate private key is in the keyring.

We can delete the public-private key pair we created for this exercise from our keyring as follows:

```{r eval=requireNamespace("gpg") && params$run_gpg}
gpg::gpg_delete(id,secret=TRUE)
```

## Data chunks {#chunks}

Data is incorporated into Rmarkdown documents using `data` chunks that consist of the data themselves preceded by a header containing a set of special chunk options describing how the data is to be processed.

### Text data chunks

The simplest possible `data` chunk is a text data chunk containing plain text. Textual data can be directly placed into a data chunk in an Rmarkdown document as follows:

````markdown
`r ''````{data output.var="d"}
a,b
1,a
2,b
3,c
`r ''````
````

```{data output.var="d",echo=FALSE}
a,b
1,a
2,b
3,c
```

When the Rmarkdown document is knitted, this chunk will put the text contents of the chunk into the variable `d`, which will then contain the chunk contents as a character string. For the example CSV data above, the character string can then be converted into a `data.frame` using `read.csv`:

```{r}
read.csv(text=d)
```

One can also load the data directly into a `data.frame` using the `loader.function` chunk option. The `loader.function` should be a function (or a character string containing the name of a function) whose first input argument will be the name of a file. A file containing the (decoded) `data` chunk contents will be passed to this function and the output will be assigned to the variable name contained in `output.var`.

````markdown
`r ''````{data output.var="d",loader.function=read.csv}
a,b
1,a
2,b
3,c
`r ''````
````

```{data output.var="d",echo=FALSE,loader.function=read.csv}
a,b
1,a
2,b
3,c
```

This will assign to `d` the output of `read.csv` applied to the CSV data in the chunk.

```{r}
d
```

Additional input arguments can be passed to `loader.function` by supplying a list as the `loader.ops` chunk option:

````markdown
`r ''````{data output.var="d",loader.function=read.csv,loader.ops=list(header=FALSE)}
a,b
1,a
2,b
3,c
`r ''````
````

```{data output.var="d",echo=FALSE,loader.function=read.csv,loader.ops=list(header=FALSE)}
a,b
1,a
2,b
3,c
```

```{r}
d
```

Note that in this case the first line of the CSV data has been treated as data instead of as a header because we supplied the `header=FALSE` optional argument.

### Text documents in data chunks

The *data* inside a text `data` chunk does not have to be scientific data. It can be any textual information, including the contents of formatting files used by Rmarkdown to generate final output documents. These include BibTeX files with references, LaTeX style files (.cls) and bibliography style files (.csl). For example, if we include the following in a Rmarkdown document:

````markdown
`r ''````{data output.file="references.bib",echo=FALSE}
@article{MeynardTestingmethodsspecies2019,
  ids = {MeynardTestingmethodsspecie,MeynardTestingmethodsspeciesinpress},
  title = {Testing Methods in Species Distribution Modelling Using Virtual Species: What Have We Learnt and What Are We Missing?},
  shorttitle = {Testing Methods in Species Distribution Modelling Using Virtual Species},
  author = {Meynard, Christine N. and Leroy, Boris and Kaplan, David M.},
  year = {2019},
  month = dec,
  volume = {42},
  pages = {2021--2036},
  issn = {0906-7590, 1600-0587},
  doi = {10.1111/ecog.04385},
  file = {/home/dmk/papers/meynard.et.al.2019.testing_methods_in_species_distribution_modelling_using_virtual_species.pdf},
  journal = {Ecography},
  keywords = {artificial species,environmental niche models,niche,simulations,species distribution modelling,virtual ecologist},
  language = {en},
  number = {12}
}

@article{SantosConsequencesdriftcarcass2018,
  title = {Consequences of Drift and Carcass Decomposition for Estimating Sea Turtle Mortality Hotspots},
  author = {Santos, Bianca S. and Kaplan, David M. and Friedrichs, Marjorie A. M. and Barco, Susan G. and Mansfield, Katherine L. and Manning, James P.},
  year = {2018},
  month = jan,
  volume = {84},
  pages = {319--336},
  issn = {1470-160X},
  doi = {10.1016/j.ecolind.2017.08.064},
  copyright = {All rights reserved},
  file = {/home/dmk/papers/santos.et.al.2018.consequences_of_drift_and_carcass_decomposition_for_estimating_sea_turtle.pdf},
  journal = {Ecological Indicators},
  keywords = {Carcass decomposition,Chesapeake bay,Conservation,Drift leeway,Drift simulations,Endangered species,Sea turtle mortality,Sea turtle strandings}
}
`r ''````
````

This will generate the file `references.bib` from the chunk contents. Note that one uses `output.file` instead of `output.var` to save the contents to a file.

As style files like the BibTeX file described above are only used by knitr/pandoc in the final formatting phase of generating an output document, these files can be generated from `data` chunks during the initial phases of the knitting. This allows them to be stored inside the Rmarkdown document itself, with no need for the external file prior to knitting.

### Base64-encoded binary data chunks

Base64 encoded binary data is incorporated into a chunk by copying the output of the `data_encode` function into the chunk and supplying the `format="binary"` chunk option:

````markdown
`r ''````{data output.var="b",format="binary",echo=FALSE}
H4sIAAAAAAAAA4vgYmBgYGZgZgNiViCTgTU0xE3XAigmDOQwAfE7KM3IwMLACaT5
kvNzCxKTS+Iz80qKUwvRZFmSEotToWK8YHEI/Q+kE2SVAwcDGNh/QKUhaph5wc6B
6GOCsFmQzWfLSS1LzSkGsgTAshBRxkQYIwnGSEbTyJqck1gM0wc3LQ3okfwiiN2o
yvMSc1NhypmwWANWzvIfzUCulMSSRL20IqBmNAM5i/LL9WCGgnzJ1AAk/v///xdk
OwBVQunahwEAAA==
`r ''````
````

```{data output.var="b",format="binary",include=FALSE}
H4sIAAAAAAAAA4vgYmBgYGZgZgNiViCTgTU0xE3XAigmDOQwAfE7KM3IwMLACaT5
kvNzCxKTS+Iz80qKUwvRZFmSEotToWK8YHEI/Q+kE2SVAwcDGNh/QKUhaph5wc6B
6GOCsFmQzWfLSS1LzSkGsgTAshBRxkQYIwnGSEbTyJqck1gM0wc3LQ3okfwiiN2o
yvMSc1NhypmwWANWzvIfzUCulMSSRL20IqBmNAM5i/LL9WCGgnzJ1AAk/v///xdk
OwBVQunahwEAAA==
```

This will place the decoded contents of the chunk into a `raw` vector `b`. Note that `format` must be specified as `'binary'` and that we have chosen `echo=FALSE` to avoid including lots of ugly encoded content in our formatted document. By default, when `format="binary"`, it is assumed that `encoding="base64"`, but this can also be supplied as a chunk option for additional clarity.

The contents of the raw vector `b` must be written to a file before they can be read back into the Rmarkdown session:

```{r}
writeBin(b,"test_output.RDS")
```

We can combine the decoding and the writing steps by specifying the `output.file` chunk option instead of `output.var`:

````markdown
`r ''````{data output.file="test_output.RDS",format="binary",echo=FALSE}
H4sIAAAAAAAAA4vgYmBgYGZgZgNiViCTgTU0xE3XAigmDOQwAfE7KM3IwMLACaT5
kvNzCxKTS+Iz80qKUwvRZFmSEotToWK8YHEI/Q+kE2SVAwcDGNh/QKUhaph5wc6B
6GOCsFmQzWfLSS1LzSkGsgTAshBRxkQYIwnGSEbTyJqck1gM0wc3LQ3okfwiiN2o
yvMSc1NhypmwWANWzvIfzUCulMSSRL20IqBmNAM5i/LL9WCGgnzJ1AAk/v///xdk
OwBVQunahwEAAA==
`r ''````
````

This will save the decoded data to the filename given by `output.file`. Then this file can be read back into the Rmarkdown session:

```{r}
readRDS("test_output.RDS")
```

Finally, we can combine all three steps (decoding, writing to disk, reading back into R) using the `loader.function` chunk option:

````markdown
`r ''````{data output.var="b",format="binary",echo=FALSE,loader.function=readRDS}
H4sIAAAAAAAAA4vgYmBgYGZgZgNiViCTgTU0xE3XAigmDOQwAfE7KM3IwMLACaT5
kvNzCxKTS+Iz80qKUwvRZFmSEotToWK8YHEI/Q+kE2SVAwcDGNh/QKUhaph5wc6B
6GOCsFmQzWfLSS1LzSkGsgTAshBRxkQYIwnGSEbTyJqck1gM0wc3LQ3okfwiiN2o
yvMSc1NhypmwWANWzvIfzUCulMSSRL20IqBmNAM5i/LL9WCGgnzJ1AAk/v///xdk
OwBVQunahwEAAA==
`r ''````
````

```{data output.var="b",format="binary",include=FALSE,loader.function=readRDS}
H4sIAAAAAAAAA4vgYmBgYGZgZgNiViCTgTU0xE3XAigmDOQwAfE7KM3IwMLACaT5
kvNzCxKTS+Iz80qKUwvRZFmSEotToWK8YHEI/Q+kE2SVAwcDGNh/QKUhaph5wc6B
6GOCsFmQzWfLSS1LzSkGsgTAshBRxkQYIwnGSEbTyJqck1gM0wc3LQ3okfwiiN2o
yvMSc1NhypmwWANWzvIfzUCulMSSRL20IqBmNAM5i/LL9WCGgnzJ1AAk/v///xdk
OwBVQunahwEAAA==
```

Given these options, during knitting, the contents of the data chunk will be decoded, written to a temporary file as binary data and then read back into the R session using the `readRDS` function. In the end, the variable whose name is given by `output.var` will be assigned the output of `readRDS`:

```{r}
b
```

### GPG-encoded data chunks

GPG chunks work similar to base64 chunks except that one must specify `encoding="gpg"`. To demonstrate this functionality, we first import into the GPG keyring the private key previously used to encode some data. We would never include the private key in an Rmarkdown document in a real use case, but this is practical for this vignette and it demonstrates another use of text `data` chunks.

````markdown
`r ''````{data key,output.file="key",include=FALSE}
-----BEGIN PGP PRIVATE KEY BLOCK-----

lQVYBF6A3p8BDADcaf7tveXZUpi0IfEpmYrPP8/OSXSh3iBkd5bdTvbq/FwLGIsD
dp/dFqAWS+0BqCIMFAtV63FUOG4kXYpkajdl2QU1Hy0aY9F9K0imc5JUM1SEry5F
CckjzDFp3u4pmmCPWKF2jVnaHzahJfKz9J9qD9BfBSynfyQU2XgsrRqNgiqeNcOi
f0674hpReawnecBwhENKMWL38O1aOtP1IDx9cFI6busiiOaIHIYYW6qbv178offy
0OWogstsQ3EJQbPBPkkgVTn8wwGUtoorc/2AonSoz99QC4nMWbBaDUGuE9O32yRv
Q7Pe6bWVBuIeV5ASAfSSEypzNHB576BF6MTy+lJvhfXI41Yu97geQJM0CplJ8xav
xAhIvrKjkDoW3zwrZlG54G2TidwEyXoDx7cyRVnCf9tsBCmhEDiKvzlg2IE9Fo65
+LWrD12qCKi7cu4XE28q4zy7S4adhUCBcuflZ8wKMVvbZRXvqnAHBAK8gQxMqHMc
EjWAb7rvmN9bkTUAEQEAAQAL/if4vPeGYaGIvhKkuSRvKOIu01O4tIMKUluF6IEX
6eVxgIuulr85CwLAMKX6fO+4+vuvwuKBARth5G+J2ygcrxE0SyJ4FejcQ0hsyg8N
lHLaoDAzyLNSc/ye8jMd75jx2yMD0rw6JBpPYMvWou4JpcNJPOOOf6ucfgGd8pI/
jjotaecpHuJgLfoapeUyqIq8JK8C/WT+EdGfCpw7YObqQq4I6ZCZPuETbKMwcQ0H
yqfWC7bK9Lk/MvbdSWDH1j70f/t1KaUEBZ2z5xTALqxaFgbwXh+7FybzV+09Sxsn
l5deeubEQXwkbPthapjRpvRo197tJRHLJ8wQVCwag39ip5cvuWQIsej3qILKTepz
VBdgZa4hIyLX8uUCAtLrVYwvWzV1oWxPLAkXJ6KPCzB0jQb7q7UUyrBOUaavdnt2
aWBz5EuXPTaMqnzWqEKIazcXqiCSNjIEv7HWcU734IGUazYper3poYgOWYYIdUes
+xbdWP/j6313N3u4a9BSd3PMvQYA4CLwr+gBfX+dybX3jq3ldB3HJS/Lv90e64rh
BarRu+ByyEO5BcVJZ+ZEUOcBjF/pvG1qI9mfqBuZX/e2aW1lmMsxcXNlWRu5b5vE
geoRwqPMNIo4JIo2hByHZeEPQLcYW/QRy5xkoNbl+udPuS3PMEUnfnPeQKursY71
ao7Zo0TUeFRemEgkvxZpFXfT+IMs9DGI/Wi6PO0ChSJ/Cu/QixgK0eJFUroNCyvl
bW+xy0GSB325wkyOM5xIny681KtvBgD7v5V6n0P2UucxZYU5hhdWaaTf5aF83vtE
o88gSU5NRO1/wPFb+AFP3fw8TNtrvRlA/OakwjL+GbfhioAJ4mtPbdGUojFIAU6X
czMHbaYyNwZTMImBW9uc2gDqta8O1HiSwC7fXnTxVoSz3E/TD6dbAnFyf1FYNntJ
PLKS9H82idCqO0nrU3LtdKJx9VHJ6wLOT16D6zZAdgNB0wK9dzStayfIqQzN/FAz
01u0ehX4SDRCxxgukdR4ZyeZJfdmC5sF+wZ/2mW4Tp7v3kutNAytk4JtMvLIhe2r
BQkYw5eUFMq7tUqXgsXMjA0pVplUSosZknCIpoyoEU7rvS9BF9xdcpRixU5kxeYY
knQg5jtb+vx3Stpp0vbuvFFaGgEJhNP6Tg3al7gBCOwEEAJmSTko4cyf1e45pIMF
+jGbIeozSjeKPWjdJCr4q05tvKgsiAe7BulgUlNhS6Ty5JyQHsiM/WZTPko2BsN2
8Apa/nuOvYwRwFLGGXVVWV3jQroPI9Hbft9ctBhUZXN0IEtleSA8dGVzdEB0ZXN0
Lm9yZz6JAdQEEwEKAD4WIQTvl6O3A7Tx/z8NeR/qzEhnRW4g7QUCXoDenwIbAwUJ
A8JnAAULCQgHAgYVCgkICwIEFgIDAQIeAQIXgAAKCRDqzEhnRW4g7WxrC/94WT6J
HEEgyb9Bskm2ik+c/qUW8w7JgizYRi6jqi8+qiIesh99MZ/XPm5mgMTIvKr0z/IG
xaU+RKYFF5DqsAc4obg/ZmClOSY9FgDWlMEm7hEqourQxfJZXGWRNcU6DTr2tC/K
GpTNkhR802LnjUePeVJU5MMuJ8eyQV+NgGhwXTIcPA6ERwHIC1n24N3QDFNoijcc
pTi5p9+N33w8fBC5ZMeZwrWI6mCJjEWVbxG2zcsIJ2t7htWRM7W1rKi5lHRpQdn/
cd9WtbdDFj7ywGPnjMB2vxYVJreENGbE/LZIZPaJKJHPReWQ+GBSGkyY7nrT32SP
R+qj5gO0Bez7F+61EDU+SXP9PJ8fyTGtUWfTsgz+fTj2TDn39y0tL1wuSciEOAjD
uia+L5qiKE9GK6mBQv78yfzZ/ZOEdJn9ZNRWs8kvs/aG9BygYMdJM5T4vvk2DcWd
m061EGTg/AVUFpMuTon9tb+RCIFfVjSzat8LWcf4Me2nJeFZu+lW/lCmxkedBVgE
XoDenwEMANPff6PrZirginP4HNK7g3ANmB3bDKCI1msAQspXMzvhtMc0Hn8DpM+r
wPUuoOo4hnYwkGHSNZ4dulrtW99mlzQWcFwDuOsvPAqc/OuEIEo0BBvc5HcpNk4d
z94Vno+Dq904VnlStf6DXpGbBFZkZBoC4XVwFUSoEjD1i967ckjFUhOxE5ynlcMb
8mpS65iml4JFd572bcuo9exJ1g7IhdgFIFoDDD2eJkxEhmEHNiVd8B9/j1GHxDCq
v/D0HNbgKuFk8WJUMYvupdqA30wAc5Ujnf+nURfNejgZTOiGXm5FZBrw/dha7yTP
/mlnNFMBKUEBrxYyPo2JVSsYfPf1WzLL1dmv8JPC5fyEKYhEC+zBvlytRWqkZV88
DumgVEdhEnnMEVlofyF8KoVMmWYA9w/FUUKiNymZlK1PEGecqliEhXh+KE03ncHh
AyEo0Zcdh5sSxUW5fNsQb+tp0fqFBs7Yye432w6ID3ZIONrnWrQ6MewWwxeAGMam
x03jgyMlCwARAQABAAv9EJ0e8iicS1JuKOfUwsWHafr26ahqlhAE2EEd+6XY06JA
PbqdhZIwk0RBjjhIz/T8vjnSqIkGQU7NdSHVqW/u/VuhFeYI0xBSIfbrckBbE9Z+
V/z7QUjPBFMcIKsLUu+dQ2yOg1b0BHAis0I3ldqrasq9CStvz4FqY8JtZFrIfGJU
rEyfYBJYEQOY/7Ne3Ap8KO/vkFx8gZLPLecgTOp2bFkCj2xbwl0rXaGl8+fP3CBA
mweyok8GGFbbVDagKE1NiukpEVzHsoMyMfPkxdIMLSj0F2GzQSnhyhyGomNstuTT
EC/i3/u7M9TRvLkpNTP3I6z5VNjayrp0NBs0z3sb1wNzrACELWbTtb/Lo5BVVD9Y
m0MQtDi8+SKzTHci2AdpvewxnhO4IiS/aXYYGcPwmEX4YdlZeV0J5mRXNsvWxYZk
HHFkbfgUkiFSFOmb9uyPD0NMldJoLXbv9+LFiU1okglietVcKK7Fyt5xCKcxbtO8
kdYJTuWonsWeyC8tz1WBBgDcq6doxs3aFSVeLcZ0//WHif+iBYlLFoexmw4irx8e
LnZilDJ5i4mwcu6Q5qxao3UEyeUC7ff//Qn846TQMDDRcC3xtrbqAqVyYBE7u9EI
OMyyCfosk8nNmVBpNdnsFm76lUyG8GiuT6b0j8BiQTRPmH4Xlh3pSiihyuTJIVhX
Y663wV8EwT9IRnYCoVqw9s5qZqJGkI4rxnABuyJui4BpmkrLry70t1xb6MdX2BPD
eK5u0YJ24AmxPW5YGvXnO0sGAPXLRfarrI9IgSz28+QpfYttOIbjp3n3AxB3ImHo
oK+CLsc1vHtsdEV8hElWo9k5EqcdlhPBbeC6IILFqT69Ldx8jK85hxR0bYs2NVLC
qyWo1T3bovPePCEenN4++VPBtVBkEt51MByNIKwC3Bw0zvHcygLcHE3iXRQ40dhq
AZWrPlOqwnC8x9+UqZoWCp/JRWD5qBjD6EPVAxwbtcUdjDOhZ1y51xbUaX59Vlul
BGLse/0Q47m71HrF+d9rGUnlQQYAkDQsdbzijmB/tVzcRXJWbZVgjwLciofxVpoM
TEYyw8+oSYDI1L3Dikejp3XymVr+9pKGmPZjLqL9Q01J9epeHt5wgLjuWTXtkVLW
kbnt7vTy257BIsHGDwiJzMI7PujTlQ4B1ZTPz2WyUJ7gn1f+J9wYpNOr7qeE2pg6
cOeiPQmT5h88jWTUH/eAJ0nAWx46kwgQY4uZz7xsFtCcwQgqVe9bD5MNv/bBUdPW
RkF8ZbRCPRk4Vl2DYM/rXC2VGCFZ6OeJAbYEGAEKACAWIQTvl6O3A7Tx/z8NeR/q
zEhnRW4g7QUCXoDenwIbDAAKCRDqzEhnRW4g7ZayC/954y+kfmjtIzSRDBRpOo2s
npOOwy7RLdOdWvab6jVecyqYsDyd/fiCXVKxALOVR31WTef00iFSLHQactwFxQyJ
zY6YO8tGkvYEXXYJR5O5MNzjlhNMndBqGIbKe9tA2BFLDD/6mmvMD/i9k+IhHzFT
NhoczB5rE9oaApMZhAj9u9Uv2zy0osfcOPcy+RN9b2noodVS/7Ei2BjWl+V/MGqa
I8oBM/ETIW/jcq+OuE8oSqoByFtFHh1DgOzOFugCWApOmAjLQwQCmDiYYtKN1GWq
l1E+txLud78ZBsJQL/78MXO9V2T2dCbcIA0vOfACuoPApfu6seRE0SLeImgoRg+8
7aX6HtiRXRjExDS26YNbGYzAvVTl3Zy1VptXOMwkh5CcIgtTcDv32pLWC3xvNydG
P4xDMM+BVuDi6QTcFfbPtqYbuuT4OFyyaSzee0oWxvKoX2pL81VnMwvb7Uy47Dxf
Ng9Af4cf3nf9UzesAVbSy1gtvlZIyX0HwtZNVLNJSS4=
=C6UF
-----END PGP PRIVATE KEY BLOCK-----
`r ''````
````

```{data key,output.file="key",include=FALSE,eval=requireNamespace("gpg") && params$run_gpg}
-----BEGIN PGP PRIVATE KEY BLOCK-----

lQVYBF6A3p8BDADcaf7tveXZUpi0IfEpmYrPP8/OSXSh3iBkd5bdTvbq/FwLGIsD
dp/dFqAWS+0BqCIMFAtV63FUOG4kXYpkajdl2QU1Hy0aY9F9K0imc5JUM1SEry5F
CckjzDFp3u4pmmCPWKF2jVnaHzahJfKz9J9qD9BfBSynfyQU2XgsrRqNgiqeNcOi
f0674hpReawnecBwhENKMWL38O1aOtP1IDx9cFI6busiiOaIHIYYW6qbv178offy
0OWogstsQ3EJQbPBPkkgVTn8wwGUtoorc/2AonSoz99QC4nMWbBaDUGuE9O32yRv
Q7Pe6bWVBuIeV5ASAfSSEypzNHB576BF6MTy+lJvhfXI41Yu97geQJM0CplJ8xav
xAhIvrKjkDoW3zwrZlG54G2TidwEyXoDx7cyRVnCf9tsBCmhEDiKvzlg2IE9Fo65
+LWrD12qCKi7cu4XE28q4zy7S4adhUCBcuflZ8wKMVvbZRXvqnAHBAK8gQxMqHMc
EjWAb7rvmN9bkTUAEQEAAQAL/if4vPeGYaGIvhKkuSRvKOIu01O4tIMKUluF6IEX
6eVxgIuulr85CwLAMKX6fO+4+vuvwuKBARth5G+J2ygcrxE0SyJ4FejcQ0hsyg8N
lHLaoDAzyLNSc/ye8jMd75jx2yMD0rw6JBpPYMvWou4JpcNJPOOOf6ucfgGd8pI/
jjotaecpHuJgLfoapeUyqIq8JK8C/WT+EdGfCpw7YObqQq4I6ZCZPuETbKMwcQ0H
yqfWC7bK9Lk/MvbdSWDH1j70f/t1KaUEBZ2z5xTALqxaFgbwXh+7FybzV+09Sxsn
l5deeubEQXwkbPthapjRpvRo197tJRHLJ8wQVCwag39ip5cvuWQIsej3qILKTepz
VBdgZa4hIyLX8uUCAtLrVYwvWzV1oWxPLAkXJ6KPCzB0jQb7q7UUyrBOUaavdnt2
aWBz5EuXPTaMqnzWqEKIazcXqiCSNjIEv7HWcU734IGUazYper3poYgOWYYIdUes
+xbdWP/j6313N3u4a9BSd3PMvQYA4CLwr+gBfX+dybX3jq3ldB3HJS/Lv90e64rh
BarRu+ByyEO5BcVJZ+ZEUOcBjF/pvG1qI9mfqBuZX/e2aW1lmMsxcXNlWRu5b5vE
geoRwqPMNIo4JIo2hByHZeEPQLcYW/QRy5xkoNbl+udPuS3PMEUnfnPeQKursY71
ao7Zo0TUeFRemEgkvxZpFXfT+IMs9DGI/Wi6PO0ChSJ/Cu/QixgK0eJFUroNCyvl
bW+xy0GSB325wkyOM5xIny681KtvBgD7v5V6n0P2UucxZYU5hhdWaaTf5aF83vtE
o88gSU5NRO1/wPFb+AFP3fw8TNtrvRlA/OakwjL+GbfhioAJ4mtPbdGUojFIAU6X
czMHbaYyNwZTMImBW9uc2gDqta8O1HiSwC7fXnTxVoSz3E/TD6dbAnFyf1FYNntJ
PLKS9H82idCqO0nrU3LtdKJx9VHJ6wLOT16D6zZAdgNB0wK9dzStayfIqQzN/FAz
01u0ehX4SDRCxxgukdR4ZyeZJfdmC5sF+wZ/2mW4Tp7v3kutNAytk4JtMvLIhe2r
BQkYw5eUFMq7tUqXgsXMjA0pVplUSosZknCIpoyoEU7rvS9BF9xdcpRixU5kxeYY
knQg5jtb+vx3Stpp0vbuvFFaGgEJhNP6Tg3al7gBCOwEEAJmSTko4cyf1e45pIMF
+jGbIeozSjeKPWjdJCr4q05tvKgsiAe7BulgUlNhS6Ty5JyQHsiM/WZTPko2BsN2
8Apa/nuOvYwRwFLGGXVVWV3jQroPI9Hbft9ctBhUZXN0IEtleSA8dGVzdEB0ZXN0
Lm9yZz6JAdQEEwEKAD4WIQTvl6O3A7Tx/z8NeR/qzEhnRW4g7QUCXoDenwIbAwUJ
A8JnAAULCQgHAgYVCgkICwIEFgIDAQIeAQIXgAAKCRDqzEhnRW4g7WxrC/94WT6J
HEEgyb9Bskm2ik+c/qUW8w7JgizYRi6jqi8+qiIesh99MZ/XPm5mgMTIvKr0z/IG
xaU+RKYFF5DqsAc4obg/ZmClOSY9FgDWlMEm7hEqourQxfJZXGWRNcU6DTr2tC/K
GpTNkhR802LnjUePeVJU5MMuJ8eyQV+NgGhwXTIcPA6ERwHIC1n24N3QDFNoijcc
pTi5p9+N33w8fBC5ZMeZwrWI6mCJjEWVbxG2zcsIJ2t7htWRM7W1rKi5lHRpQdn/
cd9WtbdDFj7ywGPnjMB2vxYVJreENGbE/LZIZPaJKJHPReWQ+GBSGkyY7nrT32SP
R+qj5gO0Bez7F+61EDU+SXP9PJ8fyTGtUWfTsgz+fTj2TDn39y0tL1wuSciEOAjD
uia+L5qiKE9GK6mBQv78yfzZ/ZOEdJn9ZNRWs8kvs/aG9BygYMdJM5T4vvk2DcWd
m061EGTg/AVUFpMuTon9tb+RCIFfVjSzat8LWcf4Me2nJeFZu+lW/lCmxkedBVgE
XoDenwEMANPff6PrZirginP4HNK7g3ANmB3bDKCI1msAQspXMzvhtMc0Hn8DpM+r
wPUuoOo4hnYwkGHSNZ4dulrtW99mlzQWcFwDuOsvPAqc/OuEIEo0BBvc5HcpNk4d
z94Vno+Dq904VnlStf6DXpGbBFZkZBoC4XVwFUSoEjD1i967ckjFUhOxE5ynlcMb
8mpS65iml4JFd572bcuo9exJ1g7IhdgFIFoDDD2eJkxEhmEHNiVd8B9/j1GHxDCq
v/D0HNbgKuFk8WJUMYvupdqA30wAc5Ujnf+nURfNejgZTOiGXm5FZBrw/dha7yTP
/mlnNFMBKUEBrxYyPo2JVSsYfPf1WzLL1dmv8JPC5fyEKYhEC+zBvlytRWqkZV88
DumgVEdhEnnMEVlofyF8KoVMmWYA9w/FUUKiNymZlK1PEGecqliEhXh+KE03ncHh
AyEo0Zcdh5sSxUW5fNsQb+tp0fqFBs7Yye432w6ID3ZIONrnWrQ6MewWwxeAGMam
x03jgyMlCwARAQABAAv9EJ0e8iicS1JuKOfUwsWHafr26ahqlhAE2EEd+6XY06JA
PbqdhZIwk0RBjjhIz/T8vjnSqIkGQU7NdSHVqW/u/VuhFeYI0xBSIfbrckBbE9Z+
V/z7QUjPBFMcIKsLUu+dQ2yOg1b0BHAis0I3ldqrasq9CStvz4FqY8JtZFrIfGJU
rEyfYBJYEQOY/7Ne3Ap8KO/vkFx8gZLPLecgTOp2bFkCj2xbwl0rXaGl8+fP3CBA
mweyok8GGFbbVDagKE1NiukpEVzHsoMyMfPkxdIMLSj0F2GzQSnhyhyGomNstuTT
EC/i3/u7M9TRvLkpNTP3I6z5VNjayrp0NBs0z3sb1wNzrACELWbTtb/Lo5BVVD9Y
m0MQtDi8+SKzTHci2AdpvewxnhO4IiS/aXYYGcPwmEX4YdlZeV0J5mRXNsvWxYZk
HHFkbfgUkiFSFOmb9uyPD0NMldJoLXbv9+LFiU1okglietVcKK7Fyt5xCKcxbtO8
kdYJTuWonsWeyC8tz1WBBgDcq6doxs3aFSVeLcZ0//WHif+iBYlLFoexmw4irx8e
LnZilDJ5i4mwcu6Q5qxao3UEyeUC7ff//Qn846TQMDDRcC3xtrbqAqVyYBE7u9EI
OMyyCfosk8nNmVBpNdnsFm76lUyG8GiuT6b0j8BiQTRPmH4Xlh3pSiihyuTJIVhX
Y663wV8EwT9IRnYCoVqw9s5qZqJGkI4rxnABuyJui4BpmkrLry70t1xb6MdX2BPD
eK5u0YJ24AmxPW5YGvXnO0sGAPXLRfarrI9IgSz28+QpfYttOIbjp3n3AxB3ImHo
oK+CLsc1vHtsdEV8hElWo9k5EqcdlhPBbeC6IILFqT69Ldx8jK85hxR0bYs2NVLC
qyWo1T3bovPePCEenN4++VPBtVBkEt51MByNIKwC3Bw0zvHcygLcHE3iXRQ40dhq
AZWrPlOqwnC8x9+UqZoWCp/JRWD5qBjD6EPVAxwbtcUdjDOhZ1y51xbUaX59Vlul
BGLse/0Q47m71HrF+d9rGUnlQQYAkDQsdbzijmB/tVzcRXJWbZVgjwLciofxVpoM
TEYyw8+oSYDI1L3Dikejp3XymVr+9pKGmPZjLqL9Q01J9epeHt5wgLjuWTXtkVLW
kbnt7vTy257BIsHGDwiJzMI7PujTlQ4B1ZTPz2WyUJ7gn1f+J9wYpNOr7qeE2pg6
cOeiPQmT5h88jWTUH/eAJ0nAWx46kwgQY4uZz7xsFtCcwQgqVe9bD5MNv/bBUdPW
RkF8ZbRCPRk4Vl2DYM/rXC2VGCFZ6OeJAbYEGAEKACAWIQTvl6O3A7Tx/z8NeR/q
zEhnRW4g7QUCXoDenwIbDAAKCRDqzEhnRW4g7ZayC/954y+kfmjtIzSRDBRpOo2s
npOOwy7RLdOdWvab6jVecyqYsDyd/fiCXVKxALOVR31WTef00iFSLHQactwFxQyJ
zY6YO8tGkvYEXXYJR5O5MNzjlhNMndBqGIbKe9tA2BFLDD/6mmvMD/i9k+IhHzFT
NhoczB5rE9oaApMZhAj9u9Uv2zy0osfcOPcy+RN9b2noodVS/7Ei2BjWl+V/MGqa
I8oBM/ETIW/jcq+OuE8oSqoByFtFHh1DgOzOFugCWApOmAjLQwQCmDiYYtKN1GWq
l1E+txLud78ZBsJQL/78MXO9V2T2dCbcIA0vOfACuoPApfu6seRE0SLeImgoRg+8
7aX6HtiRXRjExDS26YNbGYzAvVTl3Zy1VptXOMwkh5CcIgtTcDv32pLWC3xvNydG
P4xDMM+BVuDi6QTcFfbPtqYbuuT4OFyyaSzee0oWxvKoX2pL81VnMwvb7Uy47Dxf
Ng9Af4cf3nf9UzesAVbSy1gtvlZIyX0HwtZNVLNJSS4=
=C6UF
-----END PGP PRIVATE KEY BLOCK-----
```

This key then needs to be imported into the keyring:

```{r importkey,eval=requireNamespace("gpg") && params$run_gpg}
gpg::gpg_import("key")
```

Now that we have this private key, we can decode some binary data encrypted using this key:

````markdown
`r ''````{data t3,format="binary",encoding="gpg",output.var="d",loader.function=readRDS}
-----BEGIN PGP MESSAGE-----
Version: GnuPG v2

hQGMA9TPonHna5j3AQv7BIPNOSR/024iE0Gj3DCo3DvLvj/oEJ29XORHBkn4nul1
+zaRV5E/K4LCKxkkAEx/+FdM72x1hV5FF5Vf0FSet1RHiOOXPuChEEzRHOubkh/U
gw44Q72d8Dp6TOJ+1KT5k/fdkVKsOZRSttL8hvxqC4nyfObF0CkIoG+Kfx+kkYqu
araVWqNtcb3/FbtT+ZC0Hip0Ws6IJ8mGOhZdRxZ3S8KUtgf/t7S3Wa75c6L1wolT
R1/WhPgcWB4epLTvHdSmv9qcu/vFXE8SmNE5MV4V2aSTRU7y9WdPW/+XzU2Et4BK
kGyzhkI6q7QzAXFOeD1sn0uaUeH9/BDwn3AJZEXkwN4qaarPpDKjZ9GVE9Gg8521
BYe7AIZwq7sfnF+v1WyxamFYpSSAiNHze00MHPWot3Db+4SpRFYIWlYlZF00HIMo
Qspb4AmIfnNo9zj0RGG7GJoyod8ZrW4RF5iOEUWtyQ5z6LzymGTSdArWOq1fDgYW
tvEgbkbdYsJA6usJ3Zxc0sBJARfA6gDCFF72nGiAoNS98zoFjtD7hznY9DBvCOoF
jzJ8kHfPQPK9/bRVuofUDP+jOoJyf8/7eB6kANNq+XrzoZL0N42zHR2n47xupqYN
GSdsRljTra8a9zWIs9k8E5/79qRvV25c/wPeysulWkzhLCDaCMMVYJvQQ0JgT2L+
/eBtDeAuqkCqiDjeEGGB0Q4Q81IOIHMxUXFPJHvWJE4eJhnRjLmuPCBDQdL0JYTS
Bc48A/eA/cbfUr+4RluY9RLcaUHRPjKT8e7X98VdnSBPGvikVpSjR3zZhPNQs7Vb
C7H5lml4B8FpRgQBFwt2ou8URLRYR82tUa/OcsByW9jxf988YZx57a1hAw==
=m4sT
-----END PGP MESSAGE-----
`r ''````
````

```{data t3,format="binary",encoding="gpg",output.var="d",loader.function=readRDS,include=FALSE,eval=requireNamespace("gpg") && params$run_gpg}
-----BEGIN PGP MESSAGE-----
Version: GnuPG v2

hQGMA9TPonHna5j3AQv7BIPNOSR/024iE0Gj3DCo3DvLvj/oEJ29XORHBkn4nul1
+zaRV5E/K4LCKxkkAEx/+FdM72x1hV5FF5Vf0FSet1RHiOOXPuChEEzRHOubkh/U
gw44Q72d8Dp6TOJ+1KT5k/fdkVKsOZRSttL8hvxqC4nyfObF0CkIoG+Kfx+kkYqu
araVWqNtcb3/FbtT+ZC0Hip0Ws6IJ8mGOhZdRxZ3S8KUtgf/t7S3Wa75c6L1wolT
R1/WhPgcWB4epLTvHdSmv9qcu/vFXE8SmNE5MV4V2aSTRU7y9WdPW/+XzU2Et4BK
kGyzhkI6q7QzAXFOeD1sn0uaUeH9/BDwn3AJZEXkwN4qaarPpDKjZ9GVE9Gg8521
BYe7AIZwq7sfnF+v1WyxamFYpSSAiNHze00MHPWot3Db+4SpRFYIWlYlZF00HIMo
Qspb4AmIfnNo9zj0RGG7GJoyod8ZrW4RF5iOEUWtyQ5z6LzymGTSdArWOq1fDgYW
tvEgbkbdYsJA6usJ3Zxc0sBJARfA6gDCFF72nGiAoNS98zoFjtD7hznY9DBvCOoF
jzJ8kHfPQPK9/bRVuofUDP+jOoJyf8/7eB6kANNq+XrzoZL0N42zHR2n47xupqYN
GSdsRljTra8a9zWIs9k8E5/79qRvV25c/wPeysulWkzhLCDaCMMVYJvQQ0JgT2L+
/eBtDeAuqkCqiDjeEGGB0Q4Q81IOIHMxUXFPJHvWJE4eJhnRjLmuPCBDQdL0JYTS
Bc48A/eA/cbfUr+4RluY9RLcaUHRPjKT8e7X98VdnSBPGvikVpSjR3zZhPNQs7Vb
C7H5lml4B8FpRgQBFwt2ou8URLRYR82tUa/OcsByW9jxf988YZx57a1hAw==
=m4sT
-----END PGP MESSAGE-----
```

```{r eval=requireNamespace("gpg") && params$run_gpg}
d
```

```{r deletekey,eval=requireNamespace("gpg") && params$run_gpg,include=FALSE}
id = gpg::gpg_list_keys("test@test.org")$id
gpg::gpg_delete(id,secret=TRUE)
```

### Helper functions for creating and inserting data chunks

For small data chunks, one can copy-paste the (encoded) data from the command line or from a file into a Rmarkdown document. For larger data chunks, this can be awkward, so `knitrdata` includes two helper functions for creating and inserting `data` chunks into Rmarkdown documents: `create_chunk` and `insert_chunk`. We can combine these functions with `data_encode` to generate the base64 `data` chunk described above. First, we use `create_chunk` to generate the chunk:

```{r include=FALSE}
rmarkdown::draft("example.Rmd","github_document","rmarkdown",
                 edit=FALSE)
```

```{r}
library(magrittr) # For pipe operator

chunk = data_encode("test.RDS","base64") %>%
  create_chunk(chunk_label="mydata",output.var="d",format="binary",echo=FALSE,
               loader.function=readRDS)

cat(chunk,sep="\n")
```

Note that with the exception to the chunk label, chunk contents and other named arguments to this function, all the other arguments are not evaluated, so they can be given exactly as they should appear in the Rmarkdown document regardless of whether the chunk options make sense in the current context.

Next, we can use `insert_chunk` to place this chunk at a given line number in
a Rmarkdown file (here taken to be `example.Rmd`):

```{r}
rmd = insert_chunk(chunk,11,rmd.file="example.Rmd")
writeLines(rmd,"example_with_data_chunk.Rmd")
```

This will insert the new `data` chunk at line 11 in the Rmarkdown document.

`knitrdata` also includes the `list_rmd_chunks` and `splice_rmd_by_chunk` functions for identifying and potentially removing or working with the chunks in a Rmarkdown document. See the documentation and examples for these functions for more details on their use.

### Data integrity checks using md5sum

Using the procedures above should assure that data inside `data` chunks do not have errors. Nevertheless, there is always the possibility that a stray keystroke would modify the data, particularly for binary data encoded as text. To protect against this, `data` chunks can have a `md5sum` chunk option that will be used to test if the decoded data matches expectations. A MD5 sum is essentially a very large number (typically encoded in hexadecimal) derived from a file's contents that has a vanishingly small probability of being equal to the the equivalent number derived from a *different* file's contents. If two files have the same MD5 sum, they are almost certainly identical.

To use the `md5sum` chunk option, one must first determine the MD5 sum of the *decoded* source data:

```{r}
tools::md5sum("test.RDS")
```

If this character string is given as the `md5sum` chunk option to a data chunk, then the MD5 sum of the *decoded* data of that chunk will be calculated and checked against this character string. If the two do not match, an error will be generated.

We can add a MD5 sum check to the `data` chunk generated in the previous section as follows:

```{r}
md5 = tools::md5sum("test.RDS")
chunk = data_encode("test.RDS","base64") %>%
  create_chunk(chunk_label="mydata",output.var="d",format="binary",
               echo=FALSE,loader.function=readRDS,
               chunk_options_string = paste0("md5sum='",md5,"'"))

cat(chunk,sep="\n")
```

`md5sum` can be used on all `data` chunk types, including text `data` chunks
that are not encoded in any special way.

### Reading a data chunk from an external file

One disadvantage of using `data` chunks is that it can make Rmarkdown files long and difficult to navigate if you use lots of data. Rstudio can help with this problem by allowing one to hide the contents of a chunk and by using navigation tools to jump between section. In addition, using `insert_chunk` to place a `data` chunk inside a document as the last step before public diffusion may avoid having to work extensively with the large file. Nevertheless, large file size can still be a problem. To facilitate the initial construction of standalone Rmarkdown documents, data chunks can be read from external files using the `external.file` chunk option. The external file must contain the *encoded* chunk contents exactly as they would appear in a `data` chunk. The intended use of this option is that large data chunks would be placed initially in external files, but the contents of these files would be placed directly in the `data` chunks before sharing the document with others. 

```{r}
writeLines(c("This is from an external file.","It has two lines."),
           "test_external.txt")
```

````markdown
`r ''````{data ext,output.var="ext",external.file="test_external.txt"}
Content will be ignored with a warning!
`r ''````
````

```{data ext,output.var="ext",external.file="test_external.txt",include=FALSE}
Content will be ignored with a warning!
```

```{r}
cat(ext)
```

## Rstudio Addins for working with data chunks

The steps described above for encoding data, creating data chunks and inserting them in Rmarkdown documents have been combined into a set of  `Shiny` applications:

* `create_data_chunk_dialog()` for creating a data chunk and returning it to the command line after the user specifies a data file and a set of options in the `Shiny` gadget
* `insert_data_chunk_dialog()`, which has the same functionality as `create_data_chunk_dialog()`, except that the data chunk that is created is inserted at the cursor location in the active source document in the Rstudio editor
* `remove_chunks_dialog()`, which shows a data table listing all chunks in the active source document in the Rstudio editor, allowing one to select a set of chunks (i.e., rows) and then delete those chunks with the click of a button

The last two of these, `insert_data_chunk_dialog()` and `remove_chunks_dialog()`, are accessible in the Rstudio Addins menu under the `knitrdata` heading with titles `Insert data chunk` and `Remove chunks`, respectively.

Before using these tools, first install all the additional packages suggested by `knitrdata`: `shiny`, `miniUI`, `DT`, `rstudioapi`.

The use of these tools is generally self explanatory. The chunk creation apps ask for the name of the data file to be incorporated into the chunk, as well as a standard set of other chunk options, such as data format and encoding and output options. The app will attempt to suggest sensible options based on the input data file, though these can be changed afterward. For example, if a binary file is selected, then the format and encoding options will be set to `binary` and `base64` respectively. GPG encoding of data using the app is supported and will prompt the user to select the key ID(s)s of the key(s) to be used to encrypt the data. By default, `md5sum` checks of all binary data chunks will be included.

The chunk removal app present a searchable list of all chunks in the source document. To eliminate a set of chunks, one just selects the corresponding rows and clicks the `Remove chunks` button.

Note that these tools do not try to determine if the active source document is a Rmarkdown document. If undesired changes are made to a document, then one can undo (`ctrl-z`) those changes afterward.

## Use cases

There are many potential scenarios where including data inside Rmarkdown documents can be useful. The most obvious is to render a document fully standalone, perhaps for creating a single document that reproduces an entire publication or report or to make life as simple as possible for collaborators that are not comfortable managing complex Rmarkdown documents requiring multiple files. To do this, one would place not only data inside chunks in the Rmarkdown document, but also ancillary files associated with the document, such as CSS, LaTeX and CSL (bibliography) style files, and BibTeX files. These latter chunks would typically recreate these external files using the `output.file` chunk option.

`knitrdata` also has uses in cases where the objective is not necessarily to render a document fully standalone. One can use it as a convenient way to input small data tables and data vectors (e.g., see [Numeric vector data](#vectdata) below) into an Rmarkdown document. For example, the `data` chunk syntax combined with CSV data may be simpler and cleaner than creating a `data.frame` or `tibble` directly in R code. Furthermore, though markdown allows one to create small tables by hand, it is difficult or impossible to reproduce the sophisticated tables that `knitr::kable` and `kableExtra` are capable of creating, and tables created in markdown may not *float* as other tables do, potentially an issue for producing scientific publications with Rmarkdown. `knitrdata` can be used as a simple way to input these tables into R, which can then be used as input to `kable` and other tools. Finally, `data` chunks provide a convenient system for making certain Rmarkdown document text conditional on parameter values or results (see [Conditional text in Rmarkdown using knitrdata](#condtext)).

One `knitrdata` feature that may at first glance seem to be of little value is the possibility to `gpg` encrypt data chunks. Why would one want to encrypt just the data when one can encrypt the entire Rmarkdown document? However, there are many cases where the data themselves are confidential (e.g., economically important data, confidential medical information), but the methods used to analyze and synthesize those data can and should be publicly available. Furthermore, it is increasingly common to share code and documents on public collaboration websites, such as [github](https://www.github.com) or [slack](https://www.slack.com), but one may not have sufficient confidence in the privacy protections of these websites to give them access to the data themselves.  

## When not to use data chunks

Though the `knitrdata` package can be a powerful tool, it can also be abused by placing very large amounts of data inside Rmarkdown files. This will make the documents very large and difficult to navigate. The options to collapse `data` chunks in Rstudio, use Rstudio's navigation tools, use `knitrdata`'s helper functions for `data` chunk creation and the use of the `external.file` chunk option all can reduce the importance of these issues, but including very large amounts of data in Rmarkdown documents is unlikely to be optimal. In particular, Rstudio currently will not open a Rmarkdown documents larger than 5 MB in size, though larger documents can still be opened by other editors and rendering from the command line using `rmarkdown::render` is always possible. Basic prudence should be used when deciding what and how much data to include in Rmarkdown documents.

## Tips & tricks {#tt}

### Numeric vector data {#vectdata}

You can use a data chunk to load numeric vector data as follows:

````markdown
`r ''````{data output.var="v",line.sep=""}
1.2,3.4,5.6,
7.8,9.0
`r ''````

`r ''````{r}
v = as.numeric(strsplit(v,",")[[1]])
v
`r ''````
````

```{data output.var="v",line.sep="",echo=FALSE}
1.2,3.4,5.6,
7.8,9.0
```

```{r echo=FALSE}
v = as.numeric(strsplit(v,",")[[1]])
v
```


### Conditional text in Rmarkdown with missing values {#condtext}

One way to create large amounts of text in an Rmarkdown document that is only included in the document if certain conditions are met is with the `asis` language engine. For example, if one create an input parameter in the YAML header of an Rmarkdown document:

```{r eval=FALSE}
params:
  cond: TRUE
```

Then one could include content in the document based on this parameter as follows:

````markdown
`r ''````{asis eval=params$cond}
# A conditional section

This **will not** be evaluated: $5+4=`r 4+5`$
`r ''````
````

This will add a section and a paragraph with one sentence. One disadvantage of this approach is that the content of an `asis` chunk is never evaluated so one cannot use inline R code. 

This limitation can be overcome with `knitrdata` by pushing the content of the chunk into a text variable with conversion specifications that can be replaced using the `sprintf` function. For example, placing the following in the document:

````markdown
`r ''````{data include=FALSE,output.var="cond_text"}
# A conditional section

This **will** be evaluated: $5+4=%d$
`r ''````

`r ''````{r results='asis',echo=FALSE,eval=params$cond}
cat(sprintf(cond_text,5+4))
`r ''````
````

Will lead to `%d` being replaced with `9`.

## Full list of data chunk options

In addition to the standard `eval` and `echo` chunk options, `knitrdata` supports the following chunk options:

```{data output.var="t",loader.function=read.csv,loader.ops=list(sep="|",check.names=FALSE),echo=FALSE}
Chunk option|Description
`format`|One of `'"text"'` or `'"binary"'`. Defaults to `'"text"'`.
`encoding`|One of `'"asis"'`, `'"base64"'` or `'"gpg"'`. Defaults to `'"asis"'` for `format='"text"'` and `'"base64"'` for `format='"binary"'`.
`decoding.ops`|A list with additional arguments for `data_decode`. Currently only useful for passing the `verify` argument to `gpg::gpg_decrypt` for `gpg` encrypted chunks.
`external.file`|A character string with the name of a file whose text contents will be used as if they were the contents of the data chunk.
`md5sum`|A character string giving the correct md5sum of the *decoded* chunk data. If supplied, the md5sum of the decoded data will be calculated and compared to the supplied value, returning an error if the two do not match.
`output.var`|A character string with the variable name to which the chunk output will be assigned. At least one of `output.var` or `output.file` must always be supplied.
`output.file`|A character string with the filename to which the chunk output will be written. At least one of `output.var` or `output.file` must always be supplied.
`loader.function`|A function that will be passed (as the first argument) the name of a file containing the (potentially decoded) contents of the data chunk.
`loader.ops`|A list of additional arguments to be passed to `loader.function`.
`line.sep`|Only used when `encoding='"asis"'`. In this cases, specifies the character string that will be used to join the lines of the data chunk before further processing or returning them. Defaults to `'"\n"'`.
`max.echo`|An integer specifying the maximum number of lines of data to echo in the final output document. Defaults to 20. If the data exceeds this length, only the first 20 lines will be shown and a final line indicating the number of ommitted lines will be added.
```

```{r knitrdatachunkops,echo=FALSE}
knitr::kable(t,caption="Full list of knitrdata chunk options.")
```

